---
title: "Strong Projection"
author: "Katja Janelle"
date: "11/22/2021"
output: html_document
---
# Project Introduction
  
  
This project was a request from my co-worker to help create graphical illustrations for a business plan he is creating. He sent me his original data file he made calculations in Excel for which I have scrubbed to be uniform and easier for data wrangling. He is investing in an alternative cryptocurrency, Strong, and has provided me with all data according to his market assumptions.  
  
He is interested in having similar graphs to this: https://www.listenmoneymatters.com/wp-content/uploads/2019/03/LMM-Screenshots-Slider-1-1.jpg  
  
But he would like one for both post and pre-tax gains for each model.  
  
***

# Loading Data and Packages and Checking the Data

```{r}
#clear environment
rm(list=ls())

#import data file
strong <- read.csv("Strong_Data.csv")

#load packages needed
library(ggplot2) #graphing package
library(dplyr)   #pipes, mutate, etc
library(scales)  #changing scales on graph
library(ggpubr)  #ggarrange
```

```{r}
#checking that the data pulled in correctly to my data object
head(strong)
```
  
```{r}
#creating data frames one for post tax and one for net his reinvestment 

posttax <- strong %>%
                summarise(Month, Low_Strong_PostTax, Current_Strong_PostTax, High_Strong_PostTax, Highest_Strong_PostTax)

net <- strong %>%
                summarise(Month, Low_Strong_Net, Current_Strong_Net, High_Strong_Net, Highest_Strong_Net)
                
```


***
  
# Creating Graph Designs
  
Since there are many options that I can give to him (yay customization of R!), I'm going to create a few for styling purposes and see what he likes best before creating a function for easy plug and play.  
  
```{r}
posttax %>%
    ggplot(aes(Month))+
    geom_line(aes(y = Low_Strong_PostTax, color = "blue")) +
    geom_line(aes(y= Current_Strong_PostTax, color = "red")) +
    #geom_line(aes(y= High_Strong_PostTax))+
    geom_line(aes(y=Highest_Strong_PostTax, color = "green")) +
    geom_line(aes(y=90000))+ #initial investment amount line like in 
    labs(y = 'Dollar Amount') +
    scale_y_continuous(labels = dollar)+
    theme_minimal()+
    theme(axis.text.y = element_text(angle = 45))

```

Plot line are far too grouped together. What does it look like with those plots separated but grouped together on one space?  
  
```{r}
low <- posttax %>%
          ggplot(aes(Month))+
          geom_line(aes(y = Low_Strong_PostTax)) +
          geom_line(aes(y=90000))+ #initial investment amount line like in 
          labs(y = 'Dollar Amount') +
          scale_y_continuous(labels = dollar)+
          theme_minimal()+
          theme(axis.text.y = element_text(angle = 45))

current <- posttax %>%
              ggplot(aes(Month))+
              geom_line(aes(y = Current_Strong_PostTax)) +
              geom_line(aes(y=90000))+ #initial investment amount line like in 
              labs(y = 'Dollar Amount') +
              scale_y_continuous(labels = dollar)+
              theme_minimal()+
              theme(axis.text.y = element_text(angle = 45))

high <- posttax %>%
              ggplot(aes(Month))+
              geom_line(aes(y = High_Strong_PostTax)) +
              geom_line(aes(y=90000))+ #initial investment amount line like in 
              labs(y = 'Dollar Amount') +
              scale_y_continuous(labels = dollar)+
              theme_minimal()+
              theme(axis.text.y = element_text(angle = 45))

highest <- posttax %>%
              ggplot(aes(Month))+
              geom_line(aes(y = Highest_Strong_PostTax)) +
              geom_line(aes(y=90000))+ #initial investment amount line like in 
              labs(y = 'Dollar Amount') +
              scale_y_continuous(labels = dollar)+
              theme_minimal()+
              theme(axis.text.y = element_text(angle = 45))


ggarrange(low, current, high, highest, 
          ncol = 2,
          nrow = 2)

```

Also can't really tell the distinction between the four groupings. Let's try zooming in on the axis to show month 15 and onward.  
  
```{r}
posttax %>%
    ggplot(aes(Month))+
    geom_line(aes(y = Low_Strong_PostTax, color = "blue")) +
    geom_line(aes(y= Current_Strong_PostTax, color = "red")) +
    geom_line(aes(y= High_Strong_PostTax, color = "purple"))+
    geom_line(aes(y=Highest_Strong_PostTax, color = "green")) +
    #geom_line(aes(y=90000))+ #initial investment amount line like in 
    labs(y = 'Dollar Amount') +
    scale_y_continuous(labels = dollar)+
    #scale_fill_discrete(name = "Estimates", labels = c("Low", "Current", "High", "Highest")) +
    xlim(30,36)+
    ylim(1000000,6000000)+
    theme_minimal()+
    theme(axis.text.y = element_text(angle = 45)) 
    


```

This still doesn't show enough variation. Let's try a different kind(s) of graphs.  

## Bar Graphs
  
```{r}
#filter to row of data with only year one (month = 12)
posttax_year1 <- posttax %>%
                    filter(Month == 12) #%>%
                    #subset(select = -c(Month))

```





