---
title: "Use Of Force - Data from July 2020"
author: "Katja Janelle"
date: "August 3, 2020"
output: html_document
---

**Update Graphs to be prettier**  
**Include outside demographics for area population**  
**pretty up the file so HTML looks nice for displaying**  
**breakdowns by month and year?**  


```{r include=FALSE}
data <- read.csv("Use_Of_Force.csv")
library(dplyr) #mutate/summarize
library(ggplot2) #graphs
library(lubridate) #time / date manipulation
library(tidyr) #separate columns
library(kableExtra) #tables 
library(data.table) #subset rownames 

```


# **Questions I'd Like to Answer:**    
* Which precinct uses force the most? 
* What are the total count of incidents by Race?  
* What are the total count of incidents by Gender?  
* What incident types are used the most?  
* And on which groups?  
* Has incidents increased over time, decreased, or stayed the same?  


***  
  
### **Which precinct uses force the most?**   
```{r echo=FALSE, message = FALSE}
data_precinct <- data %>% 
                    group_by(Precinct) %>%
                    summarize(Total = n()) %>%
                    mutate('Percentage' = (Total / sum(Total))*100) %>%
                    mutate('Percentage' = round(Percentage, digits = 2)) %>%
                    arrange(desc(Total)) %>%
                    kable() %>%
                    kable_styling(bootstrap_options = c("striped", "condensed", "hover", "bordered")) 

data_precinct

```

**There is some bad data in this dataset.... going to remove those records from further analysis and re-run**     


```{r include=FALSE}

data <- data %>%
            subset(Precinct == 'East' | Precinct == 'North' | Precinct == 'South' | Precinct == 'Southwest' | Precinct == 'West')

```
  
  
```{r echo=FALSE, message=FALSE}
data_precinct <- data %>% 
                    group_by(Precinct) %>%
                    summarize(Total = n()) %>%
                    mutate('Percentage' = (Total / sum(Total))*100) %>%
                    mutate('Percentage' = round(Percentage, digits = 2)) %>%
                    arrange(desc(Total)) %>% 
                    kable() %>%
                    kable_styling(bootstrap_options = c("striped", "condensed", "hover", "bordered"))

data_precinct

```
  
West Precinct has the most cases with 3198 or 30% of all use of force cases, closely followed by the North Precinct with 2808 or 27% of all use of force cases. The East and South Precincts have about 2000 cases (~19% each).  

**What areas represent each of the precincts?**  

Data taken from SPD website: https://www.seattle.gov/police/about-us/police-locations/precinct-locator  

Precinct | Neighborhood
---------|------------------------------------------------------
North  |  Ballard to U-District & Above  
West  |  Downtown / Queen Anne  
East  |  Capitol Hill / Central District  
South | Columbia City / Beacon Hill ("Southeast Precinct" on website)  
Southwest  | West Seattle 
  
  
West Precinct makes sense to have the most with Downtown being heavily patrolled. Is West Seattle less crime-ridden or is West Seattle less patrolled?  

***  

### **What are the total count of incidents by Race?**  
```{r echo= FALSE,message=FALSE}
data_race <- data %>%
               group_by(Subject_Race) %>%
               rename('Subject Race' = Subject_Race) %>%
               summarize(Total = n()) %>%
               mutate('Percentage' = (Total / sum(Total))*100) %>%
               mutate('Percentage' = round(Percentage, digits = 2)) %>%
               arrange(desc(Total)) %>%
               kable() %>%
               kable_styling(bootstrap_options = c("striped", "condensed", "hover", "bordered"))

data_race

```

The most use of force cases were on White subjects with 4602 incidents in the data set or 40% of total incidents. Black subjects were most next with 3694 incidents or 32% of incidents. And then next most was unspecified racial group with 2073.  

### **What are the total populations of each Race in Seattle?**  

City of Seattle data on Demographics in Seattle as of 2018.   
Link (https://www.seattle.gov/opcd/population-and-demographics/about-seattle#raceethnicity)  
  
  
Race | Percentage
-----------|--------------------
White | 64.5% 
Black or African American | 6.8% 
American Indian & Alaska Native | 0.5% 
Asian | 14.9% 
Native Hawaiian & Other Pacific Islander | 0.3% 
Other race | 0.3% 
Two or more races | 6.0% 
Hispanic or Latino ethnicity (of any race) | 6.6% 


***  

### What are the total count of incidents by Gender?  
```{r echo=FALSE, message=FALSE}
data_gender<- data %>%
                group_by(Subject_Gender) %>%
                rename('Subject Gender' = Subject_Gender) %>%
                summarize(Total = n()) %>%
                mutate('Percentage' = (Total / sum(Total))*100) %>%
                mutate('Percentage' = round(Percentage, digits = 2)) %>%
                arrange(desc(Total)) %>%
                kable() %>%
                kable_styling(bootstrap_options = c("striped", "condensed", "hover", "bordered"))

data_gender

```

Men were subjected to use of force than Female subjects.   

**Are men arrested more than women?**  

***   
  
  
### **What incident types are used the most?**   
```{r echo=FALSE, message= FALSE}
data_force <- data %>%
                group_by(Incident_Type) %>%
                rename('Incident Type' = Incident_Type) %>%
                summarize(Total = n()) %>%
                mutate('Percentage' = (Total / sum(Total))*100) %>%
                mutate('Percentage' = round(Percentage, digits = 2)) %>%
                arrange(desc(Total)) %>%
                kable() %>%
                kable_styling(bootstrap_options = c("striped", "condensed", "hover", "bordered"))

data_force

```

Level 1 force is used the most with 8077 cases or 76% of total incidents. Level 2 follows next with 2420 cases of 23% of total incidents.  

**What does each type mean for force??**  

From: http://www.seattle.gov/police-manual/title-8---use-of-force/8050---use-of-force-definitions   

Type    | Meaning  
--------|-----------------------------  
Level 1 | Force that causes transitory pain or the complaint of transitory pain. e.g. Hobble Restraint, Pointing Gun at individual, blast balls  
Level 2 | Force that causes or is reasonably expected to cause physical injury greater than transitory pain but less than great or substantial bodily harm. e.g. taser,  OC spray, Impact Weapon, canine attack  
Level 3 - OIS | Office Involved Shooting (googled abbreviation)  
Level 3 - Use of Force | Force that causes or is reasonably expected to cause, great bodily harm, substantial bodily harm, loss of consciousness, or death.  e.g. impact weapon strike to the head 

***   

### **What incident types are used the most and on which groups?**    

```{r include=FALSE}
level1 <- data %>%
              group_by(Incident_Type, Subject_Race) %>%
              summarise(Total = n()) 
level1 <- which(level1$Incident_Type == "Level 1 - Use of Force")
            

```

```{r echo=FALSE, message=FALSE}
data_force_race <- data %>%
            group_by(Incident_Type, Subject_Race) %>%
            rename('Incident Type' = Incident_Type,
                   'Subject Race' = Subject_Race) %>%
            summarize(Total = n()) %>%
            ungroup() %>% #un-group the data so that Percentage is based off the total table output
            mutate('Percentage' = (Total / sum(Total))*100) %>%
            mutate('Percentage' = round(Percentage, digits = 2)) %>%
            arrange(desc(Total)) 

color.me1 <- which(data_force_race$`Incident Type` == 'Level 1 - Use of Force')
color.me2 <- which(data_force_race$`Incident Type` == 'Level 2 - Use of Force')
color.me3 <- which(data_force_race$`Incident Type` == 'Level 3 - Use of Force')
color.me4 <- which(data_force_race$`Incident Type` == 'Level 3 - OIS')

data_force_race <- data_force_race %>%
                        kable() %>%
                        kable_styling(bootstrap_options = c("striped", "condensed", 
                                                            "hover", "bordered"))  %>%
                        row_spec(color.me1, bold = T, color = "black", background = "#e0b1cb")%>%
                        row_spec(color.me2, bold = T, color = "white", background = "#9f86c0") %>%
                        row_spec(color.me3, bold = T, color = "white", background = "#5e548e") %>%
                        row_spec(color.me4, bold = T, color = "white", background = "#231942") %>%
                        row_spec(0, bold = 1, color = "white", background = 'black') %>%
                        scroll_box(height = "500px")
#be95c4

data_force_race
```

Level 1 and 2 use of force used the most frequently and on Black and White individuals (makes sense with the two previous statements).   
  
**Need to look at proportion of population is each group to compare to the percentage of use of force against each group.**  

***  
  
### **Has incidents increased over time, decreased, or stayed the same?**    

```{r include=FALSE}
#Need to change Occured_date_time into date / time format
data$Occured_date_time <- parse_date_time(data$Occured_date_time, "mdY IMs p")

#split date into their own columns so I can group by year, month, etc.
data <- separate(data, Occured_date_time, c("Year", "Month", "Day")) #dropping time
```

**Group by Year**
```{r include=FALSE}
graph_year <- data %>%
                group_by(Year) %>%
                count()
```

```{r echo=FALSE}

graph_year %>%
    ggplot(aes(x = Year, y = n)) +
    geom_line(group = 1) +
    geom_point() +
    geom_text(aes(label = n), vjust = 4) +
    labs(y = "# of Incidents", x ="Year")

```

Since 2020 is not finished yet, it is hard to evaluate on current year but it does appear that within the last two years that the trend is decreasing (according to the data set).  
  
Is there a seasonality trend on months that are more likely to have use of force incidents?  

***  

### **Is there a seasonality trend where which month has the most incidents?**    

**Group by Month**  

Going to exclude 2020 data in this since there is only data to beginning of July and I want to make sure it is more accurate to compare each month.    


```{r include=FALSE}
graph_month <- data %>%
                subset(data$Year != "2020") %>%
                group_by(Month) %>%
                count()
```

```{r echo=FALSE}

graph_month %>%
    ggplot(aes(x = Month, y = n)) +
    geom_line(group = 1) +
    geom_point() +
    geom_text(aes(label = n), vjust = 5) +
    labs(y = "# of Incidents", x ="Month")

```

The data depicts a seasonal trend in SPD use of force incidents in the spring to fall months (April to September). There probably is an uptick in police calls that coincide with this.  


