---
title: "Titanic Project"
author: "Katja Janelle"
date: "September 16th, 2021"
output: html_notebook
---

This is the Titanic project from Kaggle for Machine Learning.  I'm using this file as a practice for analyzing datasets as well as practicing statistical analysis methods.  

```{r}
#import the training data set
data <- read.csv("train.csv")

#load packages 
```


```{r}
#review the structure of the data 
str(data)
```
*note from Kaggle that Sibsp = the number of siblings or spouse on the titanic and Parch = number of parents or children on the titanic* 

Since there are NAs I want to remove any rows without complete information. I won't be able to later use those rows in regression analysis, etc.  

```{r}
#remove NAs 
titanic <- na.omit(data)

#how many rows were removed?
(nrow(data)-nrow(titanic))
```
177 rows were removed for missing records from the original training set.  

```{r}
#view the first few rows of the training set
head(titanic)
```

```{r}
#find the average of variables with numerical value
mean(titanic$Age)
mean(titanic$Fare)
```
Average age of the passengers was very young at 29. Average fare for the Titanic was $34.69.  
  
  

